The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-13 talk/vmil2012/paper.tex
add acknowledgements
missing word
Address some of the open tasks in the introduction and abstract sections

\begin{abstract}
Guards operations occur frequently in traces generated by tracing just-in-time
(JIT) compilers. Therefore it is important to design and implement them
carefully to find the right trade-off between execution speed, deoptimization,
and memory overhead.  In this paper we describe the design decisions about
guards taken in the implementation of the RPython tracing JIT. Furthermore we
measure various properties of guards.
% \o/
\end{abstract}

@@

\todo{the introduction needs some work}
\cfbolz{the first two two paragraphs talk about deoptimization, then it
switches to guards. I would say we should only talk about guards in the
beginning}
\bivab{Introduce in a sentence what guards are}
In this paper we describe and analyze how guards, as a concept of tracing just-in-time compilers, work. Explicating what concepts are used in the
intermediate and low-level representation of the JIT instructions and how these
are implemented.

The goal of this paper is to understand the design constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points. They need to store
enough information to rebuild the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of RPython's
tracing JIT and clarify the rationale behind the design decisions based on the
numbers provided here.

\cfbolz{this paragraph now suddenly \emph{introduces} guards, despite having talked about them already}
The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section \ref{sec:Resume Data}). During the recording phase special operations,
referred to as \texttt{guards}, are inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after the
optimization pass over the traced and later compiled parts of the benchmarks,
making guards one of the most common types of operations. Many of these guards
fail rarely or not all during execution.

@@

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind and the implementation of guards in
RPython's tracing just-in-time compiler. the contributions of this paper are:
\begin{itemize}
  \item An analysis and benchmark of guards in the context of RPython's tracing Jit.
  %An analysis of guards in the context of RPython's tracing JIT to
  %substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level parts of the JIT and describe the rationale behind the design.
\end{itemize}

@@

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend\bivab{better term for this?} related to recording and storing the
information required to restore the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

\section*{Acknowledgements}
\section*{Appendix}
\todo{remove this section and the figures}
\begin{figure*}
    \include{figures/ops_count_table}
    \caption{Relative numbers of operations in the traces generated for
    different benchmarks}
    \label{fig:ops_count}
\end{figure*}
\begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/ops_pie.pdf}
\caption{Relative frequency of operations before and after optimization}
\label{fig:ops_pie}
\end{figure*}
\bibliographystyle{abbrv}
\bibliography{zotero,paper}
\listoftodos
\end{document}

>>>>

\begin{abstract}
Tracing just-in-time (JIT) compilers record linear control flow paths,
inserting operations called guards at points of possible divergence. These
operations occur frequently generated traces and therefore it is important to
design and implement them carefully to find the right trade-off between
execution speed, deoptimization,
and memory overhead.  In this paper we describe the design decisions about
guards taken in the implementation of the RPython tracing JIT. Furthermore we
measure various properties of guards.
% \o/
\end{abstract}

@@

\todo{the introduction needs some work}
\cfbolz{the first two two paragraphs talk about deoptimization, then it
switches to guards. I would say we should only talk about guards in the
beginning}
Tracing just-in-time (JIT) compilers record and compile linear control flow paths of operations executed by an interpreter
inserting operations called guards at points of possible divergence.
In this paper we describe and analyze how guards work, explaining what concepts
are used in the intermediate and low-level representation of the JIT
instructions and how these are implemented.

The goal of this paper is to understand the design constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points, meaning the recorded
and compiled path has to be left returning control to the interpreter. They need to
store
enough information to rebuild the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of this particular
tracing JIT and clarify the rationale behind the design decisions based on the
numbers provided here.

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section \ref{sec:Resume Data}). During the recording phase \texttt{guards} are
inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after
optimizing the traces generated for the different benchmarks used in this paper.
This makes guards one of the most common types of operations. As this paper will show, many of these guards
fail rarely or not all during execution.

@@

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind the implementation of guards in
RPython's tracing just-in-time compiler. the contributions of this paper are:
\begin{itemize}
  \item An analysis and benchmark of guards in the context of RPython's tracing JIT.
  %An analysis of guards in the context of RPython's tracing JIT to
  %substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of the JIT and describe the rationale behind the design.
\end{itemize}

@@

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend\bivab{better term for this?} related to recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

\section*{Acknowledgements}
We would like to thank David Edelsohn and Stephan Zalewski for their helpful
feedback and valuable comments while writing this paper.

\section*{Appendix}
\todo{remove this section and the figures}
\begin{figure*}
    \include{figures/ops_count_table}
    \caption{Relative numbers of operations in the traces generated for
    different benchmarks}
    \label{fig:ops_count}
\end{figure*}
\begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/ops_pie.pdf}
\caption{Relative frequency of operations before and after optimization}
\label{fig:ops_pie}
\end{figure*}
\bibliographystyle{abbrv}
\bibliography{zotero,paper}
\listoftodos
\end{document}
